---
title: "Pitchfork Score Prediction"
author: "James Kowalik"
date: "14/10/2021"
output: html_document
---

## Introduction



## Cleaning the Data

For the various needs of the prediction model, during the analysis process I discovered multiple ways that the original data needed reshaping. I will remove some unnecessary variables and show what the data looks like.

```{r, eval=FALSE}
URL <- "https://dl.dropboxusercontent.com/s/cqf7cgxh91eoeyn/pitchfork.csv"
if(!file.exists("./pitchfork.csv")){download.file(URL, "./pitchfork.csv")}
```
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
p4kdata <- read.csv("./pitchfork.csv")
p4kdata <- p4kdata %>% select(-link, -review)
head(p4kdata)
```

Here I change data classes and formats, reorder all of the data by date of review and fix some formatting issues.

```{r, eval=FALSE}
p4kdata$artist <- as.character(p4kdata$artist)
p4kdata$date <- as.Date(p4kdata$date, format = "%B %d %Y")
p4kdata <- p4kdata[order(desc(p4kdata$date)),]
p4kdata$author <- p4kdata$author %>% 
  gsub("\\s+|\\.|-|é", "", .) 
p4kdata$role <- p4kdata$role %>% 
  gsub("\\s$", "", .) %>%
  as.factor(.)
p4kdata$genre[p4kdata$genre==""] <- NA
```

For the `label` variable, a variety of formatting issues arose when creating more variables that use `label`. This meant having to change punctuation, make use of spaces and deal with specific issues, namely one letter record labels and albums that were co-released on two labels.

```{r, eval=FALSE}
p4kdata$label <- gsub(" ", "", p4kdata$label)
p4kdata$label <- gsub(",", " ", p4kdata$label)
p4kdata$label <- tolower(p4kdata$label)
for(x in 1:length(p4kdata$label)){
  p4kdata$label[x] <- paste(unique(strsplit(p4kdata$label[x], " ")[[1]]), collapse = " ")
}
p4kdata$label[grep("^$", p4kdata$label)] <- "Unspecified"

p4kdata[grep("^k$", p4kdata$label),]$label <-"krecords"
p4kdata[grep("^a$", p4kdata$label),]$label <-"arecords"
p4kdata$label <- gsub("^k ", "krecords ", p4kdata$label)

p4kdata[grep("father/daughter", p4kdata$label),]$label <-"fatherdaughter"
p4kdata[grep("ever/never", p4kdata$label),]$label <-"evernever"
p4kdata[grep("i/am/me", p4kdata$label),]$label <-"iamme"
p4kdata[grep("bar/none", p4kdata$label),]$label <-"barnone"
p4kdata[grep("local331/3", p4kdata$label),]$label <-"local3313"
p4kdata[grep("sciona/v", p4kdata$label),]$label <-"scionav"
p4kdata[grep("^feel$", p4kdata$label),]$label <-"feelrecords"
p4kdata[grep("20/20/20", p4kdata$label),]$label <-"202020"
p4kdata[grep("fe/hardboiled", p4kdata$label),]$label <-"fehardboiled"
p4kdata[grep("anti-", p4kdata$label),]$label <-"antiminus"

p4kdata$label <- gsub("/", " ", p4kdata$label)
p4kdata$label <- gsub("[^0-9a-z[[:space:]]]", "", p4kdata$label)
p4kdata$label <- gsub("self-released ", "", p4kdata$label)
p4kdata$label <- gsub(" self-released", "", p4kdata$label)
```

For the review authors, there were some spelling errors detected that had to be corrected. Finally, in the last line of code, I also assigned artists with no names to `NA`.

```{r, eval=FALSE}
p4kdata$author[p4kdata$author=="DrAndyBeta"] <- "AndyBeta"
p4kdata$author[p4kdata$author=="DrewGaerig"] <- "AndrewGaerig"
p4kdata$author[p4kdata$author=="StephenMDeusner"] <- "StephenDeusner"
p4kdata$author[p4kdata$author=="StephenMDuesner"] <- "StephenDeusner"
p4kdata$author[p4kdata$author=="MarkRichardSan"] <- "MarkRichardson"
p4kdata$author[p4kdata$author=="SavyReyesKulkarni"] <- "SabyReyesKulkarni"
p4kdata$author[p4kdata$author=="PaulAThompson"] <- "PaulThompson"
p4kdata$author[p4kdata$author=="SeanFennessy"] <- "SeanFennessey"
p4kdata$author[p4kdata$author=="AlexLindhart"] <- "AlexLinhardt"
p4kdata$author[p4kdata$author=="CoryDByrom"] <- "CoryByrom"
p4kdata$author[p4kdata$author=="MarcusJMoore"] <- "MarcusMoore"
p4kdata$author[p4kdata$author=="JeremyDLarson"] <- "JeremyLarson"

p4kdata$artist[p4kdata$artist==""] <- "NA"
```

## Writing Functions

Next I look to creating new variables that will, potentially, become predictors in my model. This will involve writing functions, using for-loops, etc. Here I create a variable with all artist's names.
```{r, eval=FALSE}
artistnames <- unique(p4kdata$artist)
```

First, I will create a variable `PrevTotal` which has the total number of previous reviews the artist has had.
```{r, eval=FALSE}
PrevTotalfn <- function(name){
  artist <<- p4kdata[p4kdata$artist==name,]
  y <<- NULL
  for(x in 1:nrow(artist)){
    z <- sum(artist$date < artist$date[x])
    y <- c(y,z)
  }
  oneartist <<- cbind(artist, PrevTotal = y)
}

p4kdata2 <- lapply(artistnames, PrevTotalfn)
p4kdata2 <- bind_rows(p4kdata2)
```

Now a variable `PrevBNMTotal` which has the total number of BNMs previously given to the artist.
```{r, eval=FALSE}
PrevBNMTotalfn <- function(name){
  artist <<- p4kdata2[p4kdata2$artist==name,]
  y <<- NULL
  for(x in 1:nrow(artist)){
    a <- artist[artist$date < artist$date[x],]
    b <- sum(a$bnm)
    y <- c(y,b)
  }
  oneartist <<- cbind(artist, PrevBNMTotal = y)
}

p4kdata3 <- lapply(artistnames, PrevBNMTotalfn)
p4kdata3 <- bind_rows(p4kdata3)
```

Now a variable `PrevScore` which is the score given to the artist's most recently reviewed album.
```{r,eval=FALSE}
PrevScorefn <- function(name){
  artist <<- p4kdata3[p4kdata3$artist==name,]
  y <<- NULL
  if(nrow(artist)==1){
    y <- NA
    oneartist <- cbind(artist, PrevScore = y)
  } else {
    for(x in 1:(nrow(artist)-1)){
      a <- artist[x+1,]
      b <- a$score
      y <- c(y,b)
    }
    y <- c(y, NA)
    oneartist <- cbind(artist, PrevScore = y)
  }
}

p4kdata4 <- lapply(artistnames, PrevScorefn)
p4kdata4 <- bind_rows(p4kdata4)
```

Now a variable `PrevScoreAvg` which is the mean of all of the artists previously received scores.
```{r, eval=FALSE}
PrevScoreAvgfn <- function(name){
  artist <<- p4kdata4[p4kdata4$artist==name,]
  y <<- NULL
  if(nrow(artist)==1){
    y <- NA
    oneartist <- cbind(artist, PrevScoreAvg = y)
  } else {
    for(x in 1:(nrow(artist)-1)){
      a <- artist[(x+1):nrow(artist),]
      b <- mean(a$score)
      y <- c(y,b)
    }
    y <- c(y, NA)
    oneartist <- cbind(artist, PrevScoreAvg = y)
  }
}

p4kdata5 <- lapply(artistnames, PrevScoreAvgfn)  
p4kdata5 <- bind_rows(p4kdata5)  
```

Now a variable `PrevAuthorSame` which indicates whether the previous review for the artist was written by the same author or not.
```{r, eval=FALSE}
PrevAuthorSamefn <- function(name){
  artist <<- p4kdata5[p4kdata5$artist==name,]
  y <<- NULL
  if(nrow(artist)==1){
    y <- FALSE
    oneartist <- cbind(artist, PrevAuthorSame = y)
  } else {
    for(x in 1:(nrow(artist)-1)){
      b <- identical(artist$author[x], artist$author[x+1])
      y <- c(y,b)
    }
    y <- c(y, FALSE)
    oneartist <- cbind(artist, PrevAuthorSame = y)
  }
}

p4kdata6 <- lapply(artistnames, PrevAuthorSamefn)  
p4kdata6 <- bind_rows(p4kdata6)  
```

Now a variable `PrevAuthorSameTotal` which indicates the number of times the same author has previously reviewd this artist.
```{r, eval=FALSE}
PrevAuthorSameTotalfn <- function(name){
  artist <<- p4kdata6[p4kdata6$artist==name,]
  y <<- NULL
  if(nrow(artist)==1){
    y <- 0
    oneartist <- cbind(artist, PrevAuthorSameTotal = y)
  } else {
    for(x in 1:(nrow(artist)-1)){
      b <- sum(artist$author[x]==artist$author[(x+1):nrow(artist)])
      y <- c(y,b)
    }
    y <- c(y, 0)
    oneartist <- cbind(artist, PrevAuthorSameTotal = y)
  }
}

p4kdata7 <- lapply(artistnames, PrevAuthorSameTotalfn)  
p4kdata7 <- bind_rows(p4kdata7)  
```

Now a variable `TimeSincePrev` which shows how many days have elapsed since the artist was last reviewed.
```{r, eval=FALSE}
TimeSincePrevfn <- function(name){
  artist <<- p4kdata7[p4kdata7$artist==name,]
  y <<- NULL
  if(nrow(artist)==1){
    y <- NA
    oneartist <- cbind(artist, TimeSincePrev = y)
  } else {
    for(x in 1:(nrow(artist)-1)){
      b <- artist$date[x]-artist$date[x+1]
      y <- c(y,b)
    }
    y <- c(y, NA)
    oneartist <- cbind(artist, TimeSincePrev = y)
  }
}

p4kdata8 <- lapply(artistnames, TimeSincePrevfn)
p4kdata8 <- bind_rows(p4kdata8)
```

Now a variable `PrevTwoScoreChange` which shows the percentage growth or decline in score over the two reviews prior.
```{r, eval=FALSE}
PrevTwoScoreChangefn <- function(name){
  artist <<- p4kdata8[p4kdata8$artist==name,]
  y <<- NULL
  if(nrow(artist) <= 2){
    y <- NA
    oneartist <- cbind(artist, PrevTwoScoreChange = y)
  } else {
    for(x in 1:(nrow(artist)-2)){
      b <- artist$score[x+1]/artist$score[x+2]
      y <- c(y,b)
    }
    y <- c(y, NA, NA)
    oneartist <- cbind(artist, PrevTwoScoreChange = y)
  }
}

p4kdata9 <- lapply(artistnames, PrevTwoScoreChangefn)
p4kdata9 <- bind_rows(p4kdata9)
```

Now a variable `PrevTwoScoreAppreciated` which indicates whether the two scores prior showed an increase or a decrease in score.
```{r, eval=FALSE}
p4kdata9 <- p4kdata9 %>% mutate(PrevTwoScoreAppreciated = PrevTwoScoreChange > 1)
```

Now variables for genre. Originally, in the data, we have a `genre` variable, however problems arise as, in many cases, more than one genre is assigned to each album. For instance, see below.
```{r}
p4kdata[104,]
```

Thus I create a separate variable for each genre indicating whether this genre, be it alone or alongside other genres, appears in the album's genre tag.
```{r, eval=FALSE}
p4kdata9 <- p4kdata9 %>% mutate(GenreElectronic = grepl("Electronic", p4kdata9$genre),
                                GenreExperimental = grepl("Experimental", p4kdata9$genre),
                                GenreFolk = grepl("Folk", p4kdata9$genre),
                                GenreCountry = grepl("Country", p4kdata9$genre),
                                GenreGlobal = grepl("Global", p4kdata9$genre),
                                GenreRock = grepl("Rock", p4kdata9$genre),
                                GenreMetal = grepl("Metal", p4kdata9$genre),
                                GenreRandB = grepl("R&B", p4kdata9$genre),
                                GenrePop = grepl("Pop", p4kdata9$genre),
                                GenreRap = grepl("Rap", p4kdata9$genre),
                                GenreJazz = grepl("Jazz", p4kdata9$genre))
```

Now, using for-loops this time, a variable `LabelAvg` showing what the average of previous scores of reviews for this label is.
```{r, eval=FALSE}
p4kdata9 <- p4kdata9 %>% add_column(LabelAvg = NA)

for(x in 1:nrow(p4kdata9)){
  datecutset <- p4kdata9 %>% slice(-x) %>% filter(date <= p4kdata9[x,]$date)
  labelnames <- strsplit(p4kdata9[x,]$label, " ")[[1]]
  a <- NULL
  for(y in 1:length(labelnames)){
    b <- datecutset[grepl(paste0("\\b", labelnames[y], "\\b"), datecutset$label),]
    a <- rbind(a, b)
  }
  a <- a[!duplicated(a),]
  p4kdata9[x,]$LabelAvg <- mean(a$score)
}
```

Now a variable `LabelPrev` which shows the score that this label's previous review received. 
```{r, eval=FALSE}
p4kdata9 <- p4kdata9 %>% add_column(LabelPrev = NA)

for(x in 1:nrow(p4kdata9)){
  datecutset <- p4kdata9 %>% slice(-x) %>% filter(date <= p4kdata9[x,]$date)
  labelnames <- strsplit(p4kdata9[x,]$label, " ")[[1]]
  a <- NULL
  for(y in 1:length(labelnames)){
    b <- datecutset[grepl(paste0("\\b", labelnames[y], "\\b"), datecutset$label),]
    a <- rbind(a, b)
  }
  a <- a[!duplicated(a),]
  a <- a %>% arrange(desc(date))
  p4kdata9[x,]$LabelPrev <- a[1,]$score
}
```

Now a variable `LabelTotal` which shows the total of how many reviews the label has previously received. 
```{r, eval=FALSE}
p4kdata9 <- p4kdata9 %>% add_column(LabelTotal = NA)

for(x in 1:nrow(p4kdata9)){
  datecutset <- p4kdata9 %>% slice(-x) %>% filter(date <= p4kdata9[x,]$date)
  labelnames <- strsplit(p4kdata9[x,]$label, " ")[[1]]
  a <- NULL
  for(y in 1:length(labelnames)){
    b <- datecutset[grepl(paste0("\\b", labelnames[y], "\\b"), datecutset$label),]
    a <- rbind(a, b)
  }
  a <- a[!duplicated(a),]
  p4kdata9[x,]$LabelTotal <- nrow(a)
}

p4kdata9[p4kdata9$label=="self-released",]$LabelTotal <- 0
```

I save as an r-object for ease of future use. 
```{r, eval=FALSE}
saveRDS(p4kdata9, "./p4knewdata.rds")
```

Finally, let's take a look at the same entry we saw early with the new variables added.
```{r}
p4knewdata <- readRDS("./p4knewdata.rds")
p4knewdata[which(p4knewdata$album==p4kdata[104,]$album),]
```
